---
title: "Data Processes Project"
author: Alejandro Basco Plaza, Miriam Zaragoza Pastor, Jonatan Ruedas Mora, Raúl Cruz
  Benita and Laura Sánchez de Rojas Huerta
date: "20/12/2019"
output: pdf_document
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

## Introduction and related work

## Exploratory data analysis

### Dataset description

Before exposing the results of the exploratory data analysis carried the data, important information and background about the dataset is introduced below:

The dataset used for this work was obtained from [Kaggle's website](https://www.kaggle.com/ "Kaggle's website"). In this page, we can find a lot of information and datasets about the air quality in Madrid.

The data in this data set has been collected from the original files provided by [Madrid Open Data.](https://datos.madrid.es/portal/site/egob/menuitem.c05c1f754a33a9fbe4b2e4b284f1a5a0/?vgnextoid=9e42c176313eb410VgnVCM1000000b205a0aRCRD&vgnextchannel=374512b9ace9f310VgnVCM100000171f5a0aRCRD&vgnextfmt=default) Decide soluciones organization processed them and uploaded to the [Kaggle's website](https://www.kaggle.com/decide-soluciones/air-quality-madrid).

Originally, Madrid Open Data get the data from 24 automatic stations around Madrid. Those stations, which are set around all the districts of the city, record information about pollution in the area. The data was generated from those stations whose pollution sensors measure air quality in Madrid city.

Regarding the observations, all csv files include twelve month data, except the last one that only has data until  May. According to this, we have the followings rows:

| CSV | Observations |
| --- | ------------ |
| Stations.csv | 24 rows |
| Madrid2001.csv | 217872 rows |
| Madrid2002.csv | 217296 rows |
| Madrid2003.csv | 243984 rows |
| Madrid2004.csv | 245496 rows |
| Madrid2005.csv | 237000 rows |
| Madrid2006.csv | 230568 rows |
| Madrid2007.csv | 225120 rows |
| Madrid2008.csv | 226392 rows |
| Madrid2009.csv | 215688 rows |
| Madrid2010.csv | 209448 rows |
| Madrid2011.csv | 209928 rows |
| Madrid2012.csv | 210720 rows |
| Madrid2013.csv | 209880 rows |
| Madrid2014.csv | 210024 rows |
| Madrid2015.csv | 210096 rows |
| Madrid2016.csv | 209496 rows |
| Madrid2017.csv | 210120 rows |
| Madrid2018.csv | 69096 rows |

We can also differentiate two domains whose columns represent different data:

On the one hand, we have Stations.csv with six columns, which contains information about the stations previously mentioned, in which the data about the pollution levels of the air is collected. On the other hand, we have Madrid20xx.csv, which contain the data about Madrid's air pollution itself. Some of these csv files have 14 columns, others 16 and others 17. We have the following columns in the csv files:

| CSV | Features |
| --- | -------- |
| Stations.csv | 6 columns |
| Madrid2001.csv | 16 columns |
| Madrid2002.csv | 16 columns |
| Madrid2003.csv | 16 columns |
| Madrid2004.csv | 17 columns |
| Madrid2005.csv | 17 columns |
| Madrid2006.csv | 17 columns |
| Madrid2007.csv | 17 columns |
| Madrid2008.csv | 17 columns |
| Madrid2009.csv | 17 columns |
| Madrid2010.csv | 17 columns |
| Madrid2011.csv | 14 columns |
| Madrid2012.csv | 14 columns |
| Madrid2013.csv | 14 columns |
| Madrid2014.csv | 14 columns |
| Madrid2015.csv | 14 columns |
| Madrid2016.csv | 14 columns |
| Madrid2017.csv | 16 columns |
| Madrid2018.csv | 16 columns |

### Data wrangling
Before getting to work with the previously described data, some steps were performed in order to make the data more manageable:

#### Selected Features

In order to perform both the exploratory data analysis and the prediction methods, we will focus on the five gases that harm humans the most when they are present, and that relate the most to pollution:
1. CO

2. O3

3. SO2

4. PMH10

5. NO2

#### Datasets unification

In order to work with the data, all the yearly datasets were unified in a single one, adding one additional feature to the existing date column that all of the datasets present: the year. Thus, after this transformation, a single dataset contained all the information about Madrid's air pollution in the last 17 years, instead of 17 different datasets (one per year). Two additional columns were added: the _only_month_  and  _only_year_ columns, wihic stand for the month and the year in which the observation was recorded, respectively.

#### Missing values treatment

As we can observe in the plot below, there is a high amount of missing values in the variables that we concern about:

```{r, creation of final dataframe, echo=FALSE,message=FALSE,warning=FALSE, include=FALSE}
library(ggplot2)
library(dplyr)
library(moments)
library(gridExtra)
library(tseries)
library(ppcor)
library(corrgram)
library(car)
library(corrplot)
library(mvoutlier)
library(ppcor)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(data.table)
library(GGally)
library(ggpubr)
library(naniar)
library(mice) #treatment of missing values
library(missForest) #for prodNA
library(reshape2)
# Load CSVs
madrid_2010 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2010.csv")
madrid_2011 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2011.csv")
madrid_2012 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2012.csv")
madrid_2013 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2013.csv")
madrid_2014 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2014.csv")
madrid_2015 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2015.csv")
madrid_2016 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2016.csv")
madrid_2017 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2017.csv")
madrid_2018 <-read_csv("data/Madrid_pollution_level_dataset/csvs_per_year/madrid_2018.csv")

station <- read.csv("data/Madrid_pollution_level_dataset/stations.csv")

# Group CSVs
madrid_list <- list(madrid_2010, madrid_2011, madrid_2012, madrid_2013,
                    madrid_2014, madrid_2015, madrid_2016, madrid_2017, madrid_2018)
madrid_18 <- as.data.frame(rbindlist(madrid_list, fill = T))

remove(madrid_2010)
remove(madrid_2011)
remove(madrid_2012)
remove(madrid_2013)
remove(madrid_2014)
remove(madrid_2015)
remove(madrid_2016)
remove(madrid_2017)
remove(madrid_2018)
remove(madrid_list)
# remove(madrid_18)

# Data Wrangling
madrid_final <- madrid_18[with(madrid_18, order(date)),]
madrid_final$only_month <-lapply(madrid_final$date, month)
madrid_final$only_year <-lapply(madrid_final$date, year)
madrid_final$date<-as.POSIXct(madrid_final$date,format = "%Y-%m-%d %H:%M:%S", tz='CET')
```

```{r, creation of final dataframe part 2, echo=FALSE,message=FALSE,warning=FALSE, include=FALSE}
library(dplyr)
selected_features <- dplyr::select(madrid_final, "date", "only_year", "only_month","station", "CO", "O_3", "PM10", "NO_2", "SO_2")
madrid_final1 <- as.data.frame(lapply(selected_features, unlist))
madrid_final1$only_year <- as.factor(madrid_final1$only_year)
madrid_final1$only_month <- as.factor(madrid_final1$only_month)
madrid_final1$station <- as.factor(madrid_final1$station)
```

```{r plot missing values for selected variables, fig.width=8, fig.height=6, echo=FALSE}
library(naniar)
library(dplyr)
gg_miss_var(dplyr::select(madrid_final1, 4:8), show_pct = T)
```

Thus, in order to deal with missing values, the mean of each one of the variables for each year, month and station is computed, so that all the missing values for each one of the columns in that same year, month and station are imputed with the corresponding mean.

### Descriptive statistics analysis

After the data wrangling step, we are ready to perform the exploratory data analysis.

To start the analysis, the distribution of the variables in our dataset in terms of center, dispersion, and linear relationships descriptive measures, is analysed below.

The plot below is useful to visualize the center measures of the five selected variables:

```{r boxplots , fig.width=8, fig.height=6, echo=FALSE, warning=FALSE}
library(reshape2)
library(gridExtra)
library(ggplot2)
melt_madrid_final_1 <- melt(madrid_final1,id.vars='station', measure.vars=c('O_3','PM10','NO_2','SO_2'))
box_plot <- ggplot(melt_madrid_final_1,aes(x=variable, y=value, color=variable)) +
  geom_boxplot()+ coord_flip()+
  scale_colour_manual(values=c("magenta3","goldenrod3","brown","lightslateblue"))+
  theme(legend.position="none")+scale_y_continuous(breaks = seq(0, 700, by = 50))+
  labs(title="Distribution of pollutants",x='Pollutants',y='ug/m3')
box_plot
```

As we can see, there are many outliers or unexpected values for all of the five variables.

## Strength of relationships


## Prediction


## Results


## Discussion and future work